\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-16 paper/conclusion.tex
paper: added todos

\section{Conclusion}

As shown in \ProjectName{}, missing values and their imputation can successfully be integrated into the relational calculus and
existing plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
learning technique, as operators in the algebra and use a simple, but effective, cost model to consider tradeoffs in 
information loss and time. In effect, user preferences for one or the other can be easily incorporated by adjusting a parameter.
Simple histogram transformations
provide incrementally updated cardinality estimates across operators in the plans, which allow us to provide more accurate cost estimates.
By taking a dynamic programming approach, we can consider a variety of operator placements
and input columns, while keeping planning tractable in real-world examples.

@@

\subsection{Future Work}
\ProjectName{} opens up multiple avenues for further work. For instance, we
could extend our minimal/maximal imputation operators to consider global
information, such as the specific columns needed in operators higher up in the
query plan.  Our optimizer uses the same machine learning algorithm in all
instances of imputation operators. Integrating multiple possible algorithms
could allow for further fine-grained imputation, and honing the time complexity and
loss of these algorithms for an iterator model database could facilitate more
intelligent query plans and consistent interpretations of $\alpha$ across
strategies. For example, algorithms that learn in an online manner could
increase the efficiency of the system.  Furthermore, a multiple imputation
strategy could be followed for queries involving certain aggregates. Finally,
the underlying database for \ProjectName{} is far from a full-featured,
production database, so a natural step is to integrate imputation planning in a
widely-adopted production-quality database. This will likely expose further
opportunities for improvement.

>>>>

\section{Conclusion}
\todo{jose: rewrite this}
As shown in \ProjectName{}, missing values and their imputation can successfully be integrated into the relational calculus and
existing plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
learning technique, as operators in the algebra and use a simple, but effective, cost model to consider tradeoffs in 
information loss and time. In effect, user preferences for one or the other can be easily incorporated by adjusting a parameter.
Simple histogram transformations
provide incrementally updated cardinality estimates across operators in the plans, which allow us to provide more accurate cost estimates.
By taking a dynamic programming approach, we can consider a variety of operator placements
and input columns, while keeping planning tractable in real-world examples.

@@

\subsection{Future Work}
\todo{Jose: rewrite this}
\ProjectName{} opens up multiple avenues for further work. For instance, we
could extend our minimal/maximal imputation operators to consider global
information, such as the specific columns needed in operators higher up in the
query plan.  Our optimizer uses the same machine learning algorithm in all
instances of imputation operators. Integrating multiple possible algorithms
could allow for further fine-grained imputation, and honing the time complexity and
loss of these algorithms for an iterator model database could facilitate more
intelligent query plans and consistent interpretations of $\alpha$ across
strategies. For example, algorithms that learn in an online manner could
increase the efficiency of the system.  Furthermore, a multiple imputation
strategy could be followed for queries involving certain aggregates. Finally,
the underlying database for \ProjectName{} is far from a full-featured,
production database, so a natural step is to integrate imputation planning in a
widely-adopted production-quality database. This will likely expose further
opportunities for improvement.

### 2017-02-16 paper/experiments.tex
paper: use full width figures for tables
paper: added todos

\section{Experiments}
\todo{we of course need to rewrite this whole section}
To evaluate the performance of our system, we generate plans for queries
for two separate survey-based data sets, showing that our system
is ideal for early dataset exploration.

@@

\subsection{Queries}
We collected a set of queries (Table~\ref{fig:queries}) that we think are both representative, in that they could reasonably be written by a user in the course of data analysis, and interesting to plan.

@@

\begin{table}
\centering
 \begin{subtable}{\linewidth}
  \input{queries-cdc}
  \caption{Queries on CDC data}
  \label{fig:queries-cdc}
 \end{subtable}
 ~
 \begin{subtable}{\linewidth}
 \input{queries-fcc}
 \caption{Queries on FCC data}
 \label{fig:queries-fcc}
 \end{subtable}
  \caption{Queries used in our experiments.}
  \label{fig:queries}
\end{table}

\begin{table}
  \centerfloat
  \input{runtimes}
    \caption{Base error, percent change in error and and running time for queries
    with different imputation levels. Base error is the root-mean-square error (RMSE) between the query run on clean
    data and the query run on dirty data without imputation. Change in error is relative to the base error.}
  \label{fig:experiments}
\end{table}

Table~\ref{fig:experiments} provides a summary of our experimental results.
We run each query in four configurations: without imputation on clean data, without imputation on dirty data, and with imputation on dirty data for $\alpha = 0$ and $\alpha = 1$.
When a query returns a relation of the same ``shape'' (same number of tuples and same schema) on both clean and dirty data, we can compute an error value.
For our experiments, we report root-mean-square error (RMSE) for the base error (i.e. the error in the dirty data without imputation with respect to the clean data) and the change
in RMSE for the imputed plans.
For imputations, we run queries using the two extreme values of $\alpha$: 0 and 1, representing a focus on runtime and on information loss, respectively.
For each query, the table displays the base error, the change in error in both imputation scenarios, and the running time in seconds.

>>>>

\section{Experiments}
To evaluate the performance of our system, we generate and execute plans for queries
for two separate survey-based data sets, showing that our system
is ideal for early dataset exploration.

@@

\subsection{Queries}
\todo{Add more queries: we want ~10 (balanced between the two datasets), explain them below}
We collected a set of queries (Table~\ref{fig:queries}) that we think are both representative, in that they could reasonably be written by a user in the course of data analysis, and interesting to plan.

@@

\begin{table*}
\centering
 \begin{subtable}{\linewidth}
  \input{queries-cdc}
  \caption{Queries on CDC data}
  \label{fig:queries-cdc}
 \end{subtable}
 ~
 \begin{subtable}{\linewidth}
 \input{queries-fcc}
 \caption{Queries on FCC data}
 \label{fig:queries-fcc}
 \end{subtable}
  \caption{Queries used in our experiments.}
  \label{fig:queries}
\end{table*}

\begin{table*}
  \centerfloat
  \input{runtimes}
    \caption{Base error, percent change in error and and running time for queries
    with different imputation levels. Base error is the root-mean-square error (RMSE) between the query run on clean
    data and the query run on dirty data without imputation. Change in error is relative to the base error.}
  \label{fig:experiments}
\end{table*}

\todo{actually run experiments and rewrite this section}
Table~\ref{fig:experiments} provides a summary of our experimental results.
We run each query in four configurations: without imputation on clean data, without imputation on dirty data, and with imputation on dirty data for $\alpha = 0$ and $\alpha = 1$.
When a query returns a relation of the same ``shape'' (same number of tuples and same schema) on both clean and dirty data, we can compute an error value.
For our experiments, we report root-mean-square error (RMSE) for the base error (i.e. the error in the dirty data without imputation with respect to the clean data) and the change
in RMSE for the imputed plans.
For imputations, we run queries using the two extreme values of $\alpha$: 0 and 1, representing a focus on runtime and on information loss, respectively.
For each query, the table displays the base error, the change in error in both imputation scenarios, and the running time in seconds.
